\secdown
\secrel{About the Author}\secdown

Joe Martin worked in the construction trades after graduating from high school,
but his real love was always building and flying radio controlled model
airplanes. When he decided to turn his hobby into a business and start his own
company making com ponents fo r the radio control industry, he had to learn
about machining and toolmaking on his own. He simply couldn't afford to hire
anyone else to set up the tool s and make the molds. He has designed and taken
to mark et numerous products and owned several companies over the years. He
began his association with Sherline Products as an importer of Australian-built
lathes in the early 1970's. Since then, Joe's company has grown to become the
sole manufacturer and worldwide distributor of Sherline machine tools.

Joe was one of the founders of the sport of Formula One model aircraft
competition as well as one of its early champions. His competitive nature seems
to find its way into whatever form of fun he pursues. He has been a winner in
sports from model airplane competition to ocean sailboat racing and, most
recently, automobile racing.

Never one to be a spectator in life, he has tried and mastered many sk ills. In
this book, he passes on to you some of his hard-won knowledge about machining.
His down-to-earth style is not highly polished. In fact, if you could say that
life has put a finish on him , it would probably be described as ground or
honed\ldots Very accurate but not slick. I think his heartfelt love of good
tools and miniature machining will be apparent to all who read this book.
Working with him these past 25 years is certainly an experience I would not have
wanted to miss.

\bigskip
--- Craig Libuse

\bigskip
\textit{Joe at speed in a 1974 vintage IndyCar at Phoenix international
Raceway.}

\secup
\secup
