\textit{Under the glass watch crystal is a 1/30 scale, 1/4-20 threaded hex boltand nut along with a machined wrench to fit it. Next to it is a full size 1/4-20bolt and nut for comparison. To learn more about a man who takes miniaturecraftsmanship to extremely small scale read the following profile on JerryKieffer.}\secdown\secrel{JERRY KIEFFER\ldots Craftsmanship to the smallest detail}\textit{(Left) Jerry's Corliss steam engine is done in 1/30th scale\ldotsdown to the smallest 1/4-20 bolt! It runs as well as it looks.}\bigskip\textit{(Above) This "banjo oiler"\ is typical of the model's detail.}\bigskipJerry Kieffer is not a professional machinist. He is a marketing representativefor a utility company in Wisconsin. In fact, less than 10 years ago he had justabout no experience with miniature machine tools at all. But he is an excellentcraftsman who is good with any kind of tool and loves to make very smallprojects, so getting his hands on precision miniature machine tools opened up awhole new world of enjoyment for him.His interests are wide ranging\ldots from repairing watches and guns to makingminiature engines and tools. Jerry delights in making incredibly small parts.Some of them can only be appreciated through a magnifying glass. If a friendmakes a steam engine in the smallest scale possible, Jerry will make a half sizemodel of his friend's model! He spends several hours each evening in his shopand the amount and quality of the work he has produced is truly amazing,especially when you consider the level of perfection to which it is done.When Joe began making Sherline tools, Jerry is exactly the customer he had inmind. He uses inexpensive but accurate tools to make projects of great detailand beauty. For anyone who asks, "are tabletop machine tools accurate enough forwhat I want to do?", we simply pull out pictures of Jerry's work. Nothing moreneeds to be said.\bigskip\textit{Jerry's shop is neat and simple. The magnifying loop on his glasses isnecessary for working on and fitting the minuscule parts he enjoys making.}\bigskip\textit{A "monkey-wrench"\ in 1/7 scale.}\bigskipProject: Stover CT 2 Hit 'n Miss Engine\bigskip\textit{This running Stover "hit 'n miss"\ engine is built in 1/6 scale and hasa 1/2"\ cylinder bore. Though it may not look it, Jerry says this is one of themost difficult projects he has completed to date.}\bigskipProject: Harley Davidson Cylinders\bigskip\textit{Anyone who knows Harleys will recognize these right away. Thesecylinders for a 1/6 scale 1947 Harley Davidson "knucklehead"\ motorcycle werebuilt to test the feasibility of a project to build a complete scale motorcyclethat will not only run, but will "sound like a Harley"\ when if does. Jerry alsoplans to build a "repeater watch"\ from scratch. His goals are as big as hisparts are small, although 1 don V have the slightest doubt that if he decides totake on a challenge, he will find a way to do it.}\bigskipProject: Miniature Spark Plugs\bigskip\textit{These working spark plugs are from top to bottom: a 1/4 scale Renz 775spark intensifier plug, a 1/6 scale Champion "33"\ plug and a 1/4 scale Championprimer plug.}\bigskipProject: Miniature nut and bolt\bigskip\textit{This is the one that really blows people away! The hex bolt has a .010"shaft. The threads are 354 T.P.I. and .0005"\ deep. A hex nut is threaded ontoit. Next to the bolt is a wrench with milled contours. Compare the bolt to thescrew heads in the lady's wristwatch to the right which are the smallestfasteners commercially available. Jerry says making the nuts and bolts is easyonce you have the taps and dies\ldots it was figuring out how to make the tapsand dies that was the hard part.}\bigskip\secrel{A Flying Pendulum Clock by Jerry Kieffer}This is a variation of the "Ignatz"\ clock. It is called a "flying pendulum"clock because the ball on the string wraps around first one post and then swingsa pivot 180$^o$ to wrap the string around the other post, then back again. It ispowered by a key-wound main spring. Although this type of clock doesn't keepvery good time (accurate to within about 10 minutes a day), it is fun to displaybecause people are always fascinated by its unique movement.As a future project I would like to produce a kit with plans and materials sothat others can make this clock. It will include a video showing Jerry Kiefferbuilding it so you can see not only the steps involved, but also how a superbcraftsman plans and executes a job.\bigskip\textit{Mainspring, barrel, key and components of the winding mechanism.}\bigskip\textit{Glass or plastic domes are a good way to protect your project. This is atraditional way to display clocks where the works are to be exposed.}\bigskip\textit{Jerry made this nicely displayed set of staking tools to make it easierto press shafts and pins into gears.}\bigskip\textit{This is one of Jerry's setups using the mill and one of his specialstaking tools to press-fit the pins which act as gear teeth on the third piniongear.}\bigskip\secrel{Chapter 1— Lathe workholding}\secdown\secrel{CHUCKS}\secdownFor holding work on a lathe you have the obvious choice of using a 3-jaw or4-jaw chuck. Sherline also offers a self-centering 4-jaw chuck and an assortmentof collets. There is a safe limit to how much these chucks can be tightenedwithout damaging them. If the "tommy bars"\ are being bent while tightening achuck and the work is still coming loose you have a problem that can't be solvedby tightening the chuck to its breaking point. A 3-jaw chuck is really designedto hold only round work. It has the advantage of being able to clamp materialthat is not perfectly round with the principle being the same as a three leggedstool. Remember, if the stock isn't perfectly round it would be impossible tohave it turn perfectly true. The method you use to clamp a part in a chuck canhave a profound effect on how true it runs.\secrel{Clamping up work in the chuck}With any clamping device, you should spin the work with your fingers as it istightened. The part should feel as though it is being held securely before thefinal tightening is done. Don't insert the work into the chuck until the jawshave been adjusted to their approximate position. This keeps the work, or toolin the case of a drill bit, from getting caught between two of the three jaws.This is especially true for the expensive chucks that have the jaws rotate asthey are tightened. I don't like them because of this fact, and the extra costto have a chuck run .001"\ or .002"\ truer just doesn't seem like a goodinvestment for the home machinist.\bigskip\textit{3-jaw, 4-jaw and tailstock chucks of various sizes for use on Sherlinetools or any tool with a 3/4-16 spindle nose thread. Other threads are availableto allow the chucks to be used on some Unimat, Sears and Cowells lathes. Arborsare also available to adapt them for use on watchmakers lathes. The 2.5"\ 3-jawchuck in the middle of the bottom row is shown with its jaws reversed forholding larger work.}\bigskip\secrel{Factors affecting chuck "runout"}The chuck should run within .003 T.I.R. (Total Indicated Runout). This is theamount of variation an indicator would read if the work were rotated with thespindle. Getting a chuck to run perfectly true is difficult to accomplishbecause so many things come into play. The threads have to be true and cleanalong with the shoulder against which the chuck seats. If a part is accidentallyripped out of the chuck during machining both the spindle and the chuck can bedamaged.\secrel{The design of Sherline 3- and 4-jaw chucks}The 3-jaw self centering chuck is the most popular of all the accessoriesavailable for the Sherline lathe. It is available in both 2.5"\ diameter and 3.1"diameter. These chucks will grip round or hexagonal work quickly since the jawsmove simultaneously to automatically center the work being held. The jaws on thechuck are designed so that the same chuck can be used for both internal andexternal gripping. Jaws are reversible for holding larger diameter work, Due tothe nature of the design of a 3-jaw chuck, it cannot be expected to runperfectly true. Even 3-jaw chucks costing five times more than the one made forthis lathe will have .002"\ to .003"\ runout. If perfect accuracy is desired in aparticular operation, the use of a 4-jaw chuck is recommended. Each jaw isadjusted independently so parts can be centered with total precision. Both a2.5"\ and 3.1"\ 4-jaw chuck are available for the Sherline lathe. A self-centering4-jaw chuck is also available.\secrel{Capacities of the 3-Jaw chuck}3-jaw chucks provide the quickest and easiest way of holding work in the lathe.The Sherline 3-jaw chuck is designed so that it can be used to clamp externallyon bar stock or internally on tube stock. The 2.5"\ chuck is designed to gripfrom 3/32"\ (2mm) to 1-3/16"\ (30mm) diameter stock with the jaws in the normalposition. The 3.1"\ chuck handles stock up to 1-1/2"\ (38mm) in diameter. Forlarger diameter work, the jaws must be reversed. (See photo on previous page.)The reversible jaws can grip to 2-1/4"\ (56.0 mm) for the 2.5"\ chuck and up to2.75"\ (70 mm) for the 3.1"\ chuck. The chucks have a .687"\ (17mm) diameterthrough hole with a 3/4-16 thread.\bigskip\textit{The Sherline 3-jaw chuck with 2 tommy bars for tightening the jaws tohold a workpiece.}\bigskipJaw Location2st2nd3rdPunch MarkStandard Jaw Identification\bigskip\textit{Three-Jaw Chuck, standard jaw locations.}\bigskip\secrel{Precautions when using a chuck}NOTE: DO NOT TURN THE LATHE SPINDLE ON UNTIL THE CHUCK IS TIGHTENED. Theacceleration of the spindle can cause the scroll to open the chuck jaws if nottightened!To prevent permanent damage, a 3-jaw chuck should only be used to hold finished,turned or drawn stock. For rough castings, etc, use the 4-jaw chuck.DO NOT OVERTIGHTEN THE CHUCK. Use only moderate pressure with the Tommy Barssupplied.\secrel{Reversing the chuck jaws on a 3-faw chuck}Always start with position "A". (See illustrations above and on next page.) Toreverse the chuck jaws, rotate the knurled scroll until the jaws can be removed.They can be easily identified by the location of the teeth to the end of thejaw. To maintain chuck accuracy, the 2nd jaw must always be inserted in the sameslot even when the jaws are reversed. This slot was identified by a punch marknext to the slot. Newer chucks are engraved with the letters A, B and C asshown. Always insert the jaws in the order and at the location shown on thedrawings. Turn the scroll counterclockwise when viewed from the face of thechuck until the outside start of the scroll thread is just ready to pass theslot for the 1st jaw. Slide the 1st jaw as far as possible into the slot. Turnthe scroll until the 1st jaw is engaged.\secup\secup\secup