\secrel{Chapter 6\ --- Cutting tools for metalworking}\secdown

\secrel{6.1— GENERAL NOTES ON CUTTING TOOLS}\secdown

\secrel{Tool Deflection}

In machining, the cutting tools are the interfaces between the machine and the
part. This is where the part is cut and shaped. Machines are built which are
capable of incredibly accurate movements, but these accurate moves can't help
the process if the cutter is deflecting by large amounts. A 1/2"\ (13\,mm) end
mill can deflect as much as .050"\ (1.25mm). The deflection is a function of the
diameter of the tool compared to its length, plus the sharpness of the tool, the
feed rate and the depth of cut all have to be considered. You must understand
this fact to successfully make machined parts. It is not a business of just
numbered movements. It is a process that requires a great deal of thought and
planning to produce good parts. The better you get at anticipating these basic
problems, the better you will become at machining metals.

\bigskip
\textit{A long, thin piece of stock will deflect away from the cutting tool.
Overcoming the shortfalls of a process or the quirks of a particular machine to
make consistently good parts is what separates a hacker from a craftsman.}
\bigskip

A machinist will use every available skill to overcome these obstacles by
"sneaking up"\ to final dimensions to make the first part good. If more identical
parts are to be made, ways can then be found to speed up the process. Common
sense should tell you that if a cutter deflects a large amount on heavy cuts, it
will still deflect a small amount on light cuts. Tool or part deflection has to
be considered in every case. Whether you are operating a machine the size of a
locomotive or as small as a Sherline, the problems remain the same, but
fortunately, the costs don't. A large part can have thousands of dollars worth
of material and labor invested before a machinist touches it. How would you like
to be the employee that has to tell his boss he had just bored a hole oversize
in a twenty-thousand dollar forging?

The home machinist will have to learn these facts by trial and error. They don't
usually have the advantage of having a skilled machinist nearby to rely on for
information. I'll never understand why someone relatively new to the trade will
resent a skilled craftsman suggesting a better method of doing the job.
Information is being passed on that could be of tremendous help to them. I
imagine they have never spent days of their own time trying to solve a problem
that could have been solved in minutes by a skilled craftsman. The best and
fastest way to teach yourself is by constantly checking the part as it is being
machined. Learn how much the machine and tools deflect as they are being used.

\secrel{Overcoming Problems on a Lathe}

Lathes are easier to deal with than mills because the operations are
straightforward. They are two dimensional machines, and diameters are easy to
measure. The problems encountered with lathes usually involve the part bending.
A perfectly aligned lathe will not cut absolutely straight unless the part is
rigid enough to support itself. The shape of the cutting tool will influence the
amount of deflection. A sharp pointed tool, with the cutting edge perpendicular
to the diameter being machined, will have less deflection force than a tool with
a large radius. Large radius lathe tools can be a loser if you are working on
parts that are small or on a worn out machine because the forces are directed
away from the headstock and towards the part. This can bend the part or load
worn out slides. On a Sherline lathe a maximum tool radius shouldn't exceed
.015". On high speed steel lathe tools, I will grind a small flat on the point
rather than a radius to improve the finish. You can't grind the small radii
required free hand, but you can grind a flat on a bench grinder which will work
almost as well. Hand ground lathe tools are described completely in the section
on "grinding your own tools"\ later in this chapter. They can be not only the
fastest method of removing metal, but also the cheapest.

\secrel{Electrical discharge machining or ``EDM''}

There is another way of cutting metal with which many home machinists are not
familiar. I want you to know about this process, because I'm sure you may be
curious about how a modem machine shop deals with hardened materials. It is
called "Electrical Discharge Machining"\ or EDM. This is a slow process that does
the seemingly impossible by using an electrode to remove metal. The electrode is
made to the opposite shape you want to produce. (For example, a square electrode
will produce a square hole.) When the electrode gets within .001"\ of the work,
small sparks jump across the gap. The temperature of these sparks is hotter than
the surface of the sun, and the metal actually disintegrates. In a sense, this
happens one molecule at a time, which is why it is a slow process. It also takes
place submerged in a special oil. The electrode, which is usually a pure form of
graphite, has been shaped to burn a pocket into the hardened steel. The
electrode also disintegrates during the process. EDM machines have controls that
can be set to cut faster at the expense of the electrode; therefore, several
identical electrodes are usually used to produce each shape. Each succeeding
electrode makes the cut more and more accurate.

\bigskip
\textit{An EDM removes metal by actually vaporizing it with a high powered
electrical discharge or spark. It is capable of great accuracy and of making
shapes that would be impossible with other methods.}
\bigskip

Debris is created as the EDM operates. This will "short out"\ the cutting action
so the part must be flushed with oil to wash away the debris even though it is
submerged in oil. Much of the skill in operating these machines is how you
accomplish this. It is much more difficult than it sounds. I use an EDM machine
to build plastic molds because the electrode is many times easier to shape than
the cavity. Molds are built with materials which can be hard to machine.

\secrel{Other special uses for EDM}

A simple form of EDM machine has been developed just to remove broken taps from
parts. The electrode is simply a tube that has a diameter which is less than the
inside diameter of the hole in which the tap has broken off. Electrolyte is
pumped down the tube and the simple EDM burns the center of the tap away. The
broken tap can then be "picked out"\ and, if done correctly, the hole will not be
damaged by the sparks. There is also a highly sophisticated EDM that cuts with a
small diameter wire which is fed through the part like a bandsaw. They are used
to make cutting tools like punches and dies and are amazingly accurate but very
expensive\ --- around \$100,000. EDM machines have changed the way parts are
designed and built. Many businesses have been started that take advantage of the
capabilities of these machines which are expensive and have a limited use. I've
even seen several designs for "homemade"\ EDM machines that will work as long as
great deal of detail isn't required.

\secrel{EDM, CNC, CAD/CAM and other acronyms of the computer age}

Today, EDM electrodes are made on CNC mills which are controlled by computer
programs that take information directly from the drawing that has also been
produced on a computer. This type of work is where the term CAD/CAM (computer
aided design/computer aided manufacturing) came from. These can be exciting
times for the machinist who leams the new rules that govern this new way of
doing things.

\secup

\secrel{6.2\ --- CUTTING TOOLS FOR BOTH THE LATHE AND MILL}\secdown

\secrel{Using Center Drills}

Center drills have been designed to drill a hole that a tailstock center will
use to support the part. They have a short straight section at the end of a
sixty-degree cutting edge. As the chart shows, numerous sizes are available to
take care of any job. However, they do have a few problems you should be aware
of. The straight section doesn't have any clearance on the sides like a drill.
Without cutting oil the chips clog up the flutes and they can twist off. The
reason they don't have clearance is to make them "find"\ the center of bar stock
in a lathe. Drills that are flexible can start drilling off center without the
help of a center drill to get them started. Even with cutting oil you can't
drill to the required depth without backing out the center drill completely and
adding cutting oil to it's tip. The straight part of the hole was to make a
pocket to keep lubricants for lubricating the supporting sixty-degree tailstock
dead center. A great deal of heat from friction can be generated at the center
without proper lubrication. White lead was the preferred lubricant of its day
because it didn't gall, but for obvious health reasons it isn't available any
more. On lathes we now use live centers which rotate with the part to eliminate
friction.

\bigskip
\textsc{COMMONLY AVAILABLE CENTER DRILL SIZES}
\bigskip

\secrel{The difference between a Center Drill and a Spotting Drill}

Center drills are an important tool in machining and are used as much on a mill
as a lathe. It is a simple fact of life that drills will "walk"\ when starting a
hole that hasn't been started with a short, rigid drill. Center drills were the
only tools available that could start holes without wandering. The only problem
was, it took too long to drill a pocket for the lubricant that wasn't needed.
Years ago, I saw this problem

\bigskip
\textit{A center drill (top) and a spotting drill (bottom).}
\bigskip

and started making "spotting drills"\ for use in our own shop. We could save
minutes of machine time on one part by not using center drills. They have
finally brought to the market drills of this type called "spotting drills". When
compared to the complexity of a standard drill, spotting drills cost more than
they should, for they could be manufactured cheaper than a center drill.

\bigskip
\textit{A center drill being used to find the center of a piece of stock. The
$60^{o}$ angle will locate the point of the tailstock center for turning between
centers. The smaller hole will serve as a reservoir for lubricant if a live
center is not to be used.}
\bigskip

A drill press can be the most dangerous machine in a machine shop. They are so
simple to use that operators don't give them the respect they deserve. New
employees are often given drill press jobs to start their machining career.
Never hold a part by hand when drilling into metal or similar types of
materials. The drill can grab the part and start it spinning at the RPM the
machine was set at. The greatest chance of this happening is when a hole is
enlarged with a larger drill. The softer the material, the better the chance of
this happening. The first part of this scenario is the where the drill screws
into the part, lifting it off the table. When the drill jams in the part, the
part will spin. If this happens. step away and try to safely shut down the
spindle. Whatever you do, don"t be a hero and try to grab the part. You could
end up looking like you have just been through a war in the 16th century. Also,
when drills are broken off they may leave a very sharp point in either the
spindle or the part that could cut to a depth that could result in permanent
injury. Clamp your parts down firmly and cut down on injuries.

\secrel{Who came up with these stupid drill sizes anyway?}

Manufacturers of cutting tools sometimes have a strange way of applying logic.
Take the standard drill indexes that are available today. They are still the
same as sets from fifty years ago. If anyone could explain the logic of the
sizes they manufacture I would love to know, They simply keep making the same
old sizes they have always made without questioning the logic. Overseas
manufactures aren't any better for they just copy what is available. (Obviously
this is one of my pet peeves.) The quality of the better drills manufactured
today is superb, but it is too bad they don't make them in the proper sizes. The
better industrial suppliers may offer high and slow helix (flute twist), but I
would recommend that you stick with the standard "jobber"\ drills. They are
cheaper and work quite well. The hobbyist/machinist will want to buy a drill set
for working with the miniature machine tools. My first choice would be a number
1 to 60 set. (Sizes .040"\ to .228".) This set will have the drills you need for
drilling and tapping small holes. A miniature tool doesn't have the power or the
size for large drills. On smaller machines, larger holes are bored rather than
drilled.

\bigskip
\textit{The metal box holds a set of drill bits sized from number 1 through 60.
The red plastic box holds a selection of very small bits from number 61 through
80. In the foreground are \#1, \#60 and \#80 sizes.}
\bigskip

There is a series of drills called "screw machine drills"\ that are shorter and
better suited for small machines. They make them in both left and right hand.
Only right hand drills will work on a Sherline because of the direction of
rotation. Use high speed steel drills because of their low cost and long life
when properly used. Stay away from carbide drills unless you have a particular
need for them. They are very brittle and most small tools can't produce the
spindle speeds needed for carbide to have an advantage. Printed circuit drilling
machines, for example, may run carbide drills at over 60,000 RPM.

\secrel{Using drill bits properly}

Drills get more abuse than any other metal cutting tool made. Watching a home
repair expert stick a drill back in a hole that they just drilled with a slight
error and try to make the hole oblong by bending the drill should give a
machinist the horrors. To begin with, the flutes on a drill were never designed
to cut, but somehow they do when enough pressure is applied. They are designed
to pull the chips out of the hole and a small amount of coolant is a must,
especially for holes going more than two drill diameters deep. If you want
accurately drilled holes, you can't abuse drills. Small diameter drills are
usually discarded when dull to maintain accuracy. Hand-sharpened drills will
normally drill oversize because the cutting lips are not exactly equal. To keep
drills sharp, they have to be turned at a speed that is forty percent or less of
the material cutting speed they are drilling. Exceeding this speed can instantly
destroy the cutting edge. Dulling tools in this manner is foolish because speed
is one of the easiest variables to control. Think about the way American Indians
would start the campfire. They would turn a straight, dry stick back and forth
with the palms of their hands until friction generated enough heat to start a
fire on the wood where the stick point was rubbing. Now consider how fast they
could have started that fire if they powered that stick with a five horsepower
motor! Hopefully I have made my point.

With the centering hole drilled and the RPM set correctly it is time to drill
the hole to size. The Sherline motor doesn't have the horsepower for large
diameters and high feed rates, but that isn't a problem as long as you work
within the capabilities of your tools. It just takes a little longer. The real
problem is keeping a constant feed rate and knowing how deep the drill can go
without "pulling"\ the drill.

The drill point must be lubricated and the chips must be removed from the drill
flutes. With automatic production machinery, the drill can go as much as three
and a half times the drill diameter into the hole before it is backed out to
clear the chips. From that point on, I usually program depth of cut in one drill
diameter increments. Using manual machines I would recommend two times the drill
diameter for the first cut and one diameter beyond that point. When using small
drills, the amount that can be drilled without clearing the drill is as small as
the drill diameter. This is very important and the reason I am emphasizing it. A
drill that has been broken off in a part will usually result in "scrap"". To
make matters worse, when you start over again, unless you have a spare bit, the
one you need is now broken. Good drilling technique makes this one problem that
is avoidable.

\secrel{Should a hole be drilled in stepped sires or directly to size?}

I have found that once the part is center drilled, drilling a hole to size with
a single drill will give better results than enlarging a smaller hole with a
series of larger drills. Every time a different tool enters a part it can pick
up off center, because the drill before it may have left a sharp burred edge in
the hole. If this method must be used, make sure the spotting or center drill
will cut to a diameter that includes all the sizes you will be using and has a
ninety or sixty degree tip. Brass is a difficult material in which to enlarge a
drilled hole. In fact, it can be downright dangerous, because a standard drill
can be pulled into a pre-drilled hole and lock up. Machinists often put a slight
flat on the cutting lip of drills used to enlarge a hole in brass to avoid this
problem.

\secrel{Drilling stainless steel}

First-time machinists will discover it's a whole new ball game the first time
they attempt to drill a hole in a piece of stainless steel. It is called "work
hardening"\ and it is just as the name implies. I'm sure everyone has taken a
wire and bent it back and forth until it broke. What was done was the metal kept
getting harder at the bend until it became so brittle it fractured. Stainless
steel as well as many other exotic metals can surface harden while being
machined. Small drills will make this fact even more apparent. If a drill is
allowed to rub the surface without cutting, it will work harden the surface it
is trying to drill. This happens at the bottom of a hole that is being drilled.
The cause can be a dull drill or not using a sufficient feed rate to keep the
drill cutting. This will work harden the surface, which will dull the drill,
which will work harden the surface even more until the drill will not cut at
all. To work around this problem, the spindle has to be turning slow enough to
have a feed rate that allows the drill tip to maintain a constant, uninterrupted
cut. The calculated spindle speed may be much faster, but the machinist can't
keep up with the higher RPM. Also, don't leave the drill at the bottom of a hole
when it isn't cutting. Raise it immediately. If a hole gets work hardened, check
the drill for sharpness and start drilling again with a slight "tap". This will
sometimes allow the drill point to break through the hardened surface and start
drilling again.

\secrel{A Note on very large drill bits}

Drill bits at the large end of the size scale usually have a tapered shank and
could never be used on a Sherline, for they are as big as the machine. However,
there is still something to be learned from large diameter drills. In many
cases, larger drills have a tapered shank with a "tang". A tang is a rectangular
shape on the small end of the taper that engages with a holder designed to
accept it so the drill cannot spin. The added cost for manufacturing this type
of drill is worth every penny if it keeps the drill from spinning, as this will
ruin the shank. The way standard twist drills are kept from spinning is by
tightening the drill chuck with a key to create enough friction to lock them in
place. If they are not tightened enough, they will spin in the chuck, and you
simply can't tighten the chuck enough by hand without a key. Many machinists
never learn this fact and the results of their carelessness can be found in tool
cribs around the world.

\bigskip
\textit{A large drill bit with a tapered shank. On larger machine tools where
cutting forces are higher the flat tang keeps the drill bit from spinning in the
chuck. Miniature machine tools don 7 have enough power to require this kind of
holding force, but you still need to tighten your chuck with a key so the drill
can 7 be ruined by spinning in the chuck.}
\bigskip

Additional information on large drills is not included because it isn't very
useful for small machines like the Sherline. \emph{Machinery's Handbook} will be
the best source of information for large drills. I have also learned much
information from catalogs on cutting tools printed by mail order suppliers. I
would also suggest you inspect a small drill under a magnifying glass some time
to appreciate the level of perfection to which these tools have been made. Their
very reasonable cost will be appreciated even more once you get a really close
look.

\secrel{Countersinks}

The countersinks I prefer to use have only one cutting lip, and they will cause
you very few problems with chatter. Countersinking tools are used to chamfer the
sharp edge left after a hole has been drilled. A wide variety of angles are
available. They are also used to allow flat-headed screws to fit flush with the
part surface. They should be run at a very low rpm when used in manual machines.
You can instantly destroy these tools in stainless steel by using excessive
speed with a hand drill. If the hand drill you are working with doesn't have a
speed control, keep the RPM low by pulsing the trigger switch. The speed should
be such that you can see the tool cutting. If the tool is not perpendicular to
the hole, the chamfer will be uneven.

\secrel{Step drills}

Step drills have one or more diameters and are made to eliminate changing tools
to save time in machining operations. Only consider the step drill's smallest
diameter for feeds and its largest diameter for speeds. It is easy to twist the
tip off if not enough cutting oil is used. They are expensive, and because of
the limited sizes available will usually be custom-made by local toolmakers.
Only use them when the cost is justified by the time saved.

\secrel{Counterbores}

Counterbores are used to enlarge existing holes. They have a "pilot"\ shaft which
can be changed to different sizes on larger diameters counterbores. One of the
most common uses for these tools today would be counterboring a clearance screw
hole for a socket head cap screw. (Many times these screws are referred to as
"Allen cap screws".) Spot facing rough or curved castings so bolts can be pulled
down on a

\bigskip
\textit{Top: a 1/4"\ short shank counterbore of high speed steel. Above: a longer
style counterbore is well suited for screw heads and spring pockets. There are
many more styles and shapes,}
\bigskip
\textit{A pilot sized to the drilled hole goes in the front of each counterbore.
They are normally sold separately and are held in place with a set screw.}
\bigskip

flat surface is another common use. If the pilot is the right size it keeps the
tool from wandering, Counterbores will usually have three flutes and a very slow
helix, making their diameters difficult to measure, They have to be turned at
low RPM with a moderate amount of pressure similar to using a countersink. They
are expensive.

\secrel{Reamers}

Reamers are available in more sizes than drills. A reamed hole will have more
accurate diameters and better finishes than a drilled hole. However, if they are
not used properly, a good drilled hole may be better than a reamed hole. They
come with straight and spiral flutes. In theory, straight flutes will work
better if the hole goes all the way through the part because a lot of chips get
in front of the cutting edge. Spiral flutes will help pull the chips from the
bottom of a "blind"\ hole. A right-hand cut, left-hand spiral will push the chips
forward and should never be used in a blind hole. I always order straight flutes
for general use. A feed rate of .001"\ per tooth is recommended, so the RPM must
be slow enough to maintain this feed. Ample cutting oil must be applied, and the
reamer must be pulled out of the hole often so chips can't build up in the
flutes. The reamer must be located directly over the center of the hole to work
properly.

The safest way is to ream the hole to size immediately after drilling. I usually
allow less than .008"\ (.2mm) on sizes less than 5/16"\ (8mm) and for over this
diameter I allow 1/64"\ (.4mm) for clean-up with a reamer. A reamer can start
cutting off-center if the drilled hole has chips lying on the hole edge; in fact
on CNC machines, we occasionally put a program stop to allow the operator time
to blow the chips out with an air hose. What happens is the chip will start
rotating with the reamer, shutting down the cutting action on one side. The
reamer will cut oversize until the chip is forced from its location. It will
then start cutting on-size leaving the hole with a "bell-mouthed"\ entrance.
Fractional inch sizes are quite a bit cheaper than decimal size reamers. Try to
design your parts to have finish diameters that take advantage of this fact.

\secrel{Taps}

Taps are cutting tools to cut screw threads in holes (see thread cutting
instructions for thread definitions) and are made in many sizes and shapes. Taps
cut several thread shapes and the sixty-degree thread is standard. The "acme"
and "square"\ threads are not available in diameters under one half inch and will
very seldom be used by the hobbyist. For the type of work done in miniature
machining, I would not consider large taps. If needed, you can usually get the
information from most industrial supply catalogs.

The same rules should be applied to large and small taps. Taps are defined by
their diameter first, which may be a wire gauge size, and followed by the number
of threads per inch. Metric taps will be defined by diameter and actual pitch.
Drill and tap charts will give the information to cut 75% of a full thread.
Engineers have discovered long ago that a full thread isn't necessary. A 75%
thread is almost as strong and the difficulty in cutting a full thread just
isn't worth the extra effort and cost.

To tap a high quality thread, you need to start with a proper hole size and use
a lubricating cutting fluid. A chamfered hole will start the tap better and will
not raise the surface where the hole is being tapped; however, thin parts may
require no chamfer in order to have the maximum number of threads for a given
thickness.

\secrel{Why cheap taps are a poor investment}

If you have a choice as to the size you can use, look at a tool catalog and pick
out a size that is readily available. The savings in cost will be worth the
effort. Remember also that just because the size may be listed doesn't mean you
can buy it at every industrial supplier, never mind a local hardware store. Buy
taps before you start the job, and it would be wise to buy an extra one. They
are very easy to break. If possible, I get my tapping done early on complex
parts. If a tap does break and ruins the part, you are not scrapping a lot of
labor. If you are having a problems tapping holes, the first question to be
asked is about the quality of the taps you are using. Inexpensive tap sets can
be so bad they wouldn't cut butter. They usually are found in department stores.
Stay away from them no matter how tempting they may look. Tap sets sold by
industrial suppliers are better, but I still advise buying only what you need,
when you need it. Buy quality taps and save hours of grief.

Taps are made in a variety of styles, and my favorite are the "spiral point gun
taps". In smaller sizes they have two flutes and are the strongest series of
taps available. "Spiral pointed"\ means that the chip is pushed ahead of the tap.
This eliminates the need to back up a tap to break the chip as four-fluted hand
taps require. These taps are made with two choices for the cutting tip: "plug"
and "bottoming". Use the plug style for general use. Spiral point taps have been
designed to push the chips out the bottom of the hole that is being tapped. In
blind holes, the chips will pile up at the bottom and could cause a problem.

\bigskip
\textsc{SPIRAL POINT TAP}

\textsc{SPIRAL FLUTE TAP}
\bigskip

\secrel{Spiral pointed taps vs. spiral fluted taps}

Don't confuse spiral pointed taps with spiral fluted taps. Spiral fluted taps
pull the chips out of the hole, which may seem like an excellent idea until the
strength of the tap is considered. They break easily, because under load they
"unwind"\ and jam in the hole. The problem eliminated by these taps can be
canceled out by their inherent weakness. If possible, drill the hole deeper than
it will be threaded to make room for these chips. After tapping, the chips can
be removed from the hole with tweezers.

A problem can arise when a tap is reversed. The chips can jam the tip in a blind
hole, breaking the tap, but tap breakage is less with two-fluted than with
four-fluted "hand taps"\ because of the extra strength of these fine cutting
tools. In small sizes, the four-fluted hand taps should be avoided but some
sizes are only available in that form. "Starting"\ taps have a shallow cutting
angle mat will cover many threads. Each cutting tooth will form a small amount
of the thread. While this may seem like a good idea, it does not always work
well. With many threads cutting at the same time, the torque required to drive
them may be higher than a plug style spiral pointed gun tap, but they do start
the tap straight down the hole with less effort.

\secrel{Tapered pipe thread taps and dies}

Pipes used in plumbing use a National Pipe Thread (NPT) standard which is a
tapered thread. Tightening the parts together seals them as the tapers meet,
making a tight joint. When tapping a hole for a tapered thread, measure the part
you will be threading into it to make sure you don't run the tap too deep. NPT
threads start at 1/16-27 and go up to over 3 inches. You might use some of the
smallest sizes on miniature plumbing fittings for steam engine models or pipe
fittings in small engine blocks.

\secrel{Hand tapping a hole}

When tapping holes by hand, the tap must be perpendicular to the hole. I
recommend having the part clamped down so the tap will be pointed towards the
center of the earth. It is easier to line up when the tap and handle are square
to its surroundings. The quickest way to break a tap is casually holding the
part in your hand and a tap in the other. If the tap is not "square"\ it will
not follow the hole and will progressively start to cut one side more than the
other. This will quickly reach the point of no return when the cut exceeds die
strength of the tap. It is very difficult to straighten a tapped hole. Again,
you must use a cutting fluid. It not only keeps chips from sticking to the
cutting edge, it also keeps the tap from sticking to the cut thread. Remember
that "like"\ materials will gall and stick together, and this is one reason taps
break at a alarming rate.

\emph{NOTE}: Broken taps can be dangerous! Pieces of the tap can be thrown at a
very high speed and easily do eye damage, Always wear eye protection. The broken
tap can leave a sharp cutting edge where it has broken and give a nasty, deep
cut that could require stitches.

\secrel{Tapping stainless sleel}

Hand tapping a small diameter screw thread of 75\% may seem impossible if the
material is stainless steel. The tip of the tool will twist' to the point of
almost breaking before the tap will even start to cut. Dull or poor quality taps
will not stand a chance in stainless. A 65\% thread may help, but a change such
as this must be authorized if the part isn't for your own use, and considered if
it is for your own use. Don't use a tap until it breaks. They are disposable
tools and have a life-span that can be short with exotic materials.

\secrel{Tap wrenches}

The tap wrenches for holding taps used by amateurs will usually be the standard
hand tap holder. Buy the size that just goes to one-quarter inch and, if you can
locate a smaller size, buy it too. It will be useful to have a very small tap
wrench for the small sizes that you will encounter while working on very small
parts. When you work with equipment that is too big for the job, you lose that
"feel"\ that keeps taps from breaking. A tap wrench has a chuck similar to a
drill chuck but it has only two jaws. The jaws allow the tap to be held by the
square, machined end on the tap. Good quality tap wrenches will clamp the tap in
line with the tap wrench body, making it easier to get the tap square with the
part. When working with full size mills, I will use the spindle to line the tap
over the hole. Some tap wrenches have a sliding shaft protruding from the
handle. Using this shaft, the holder can then be held in a collet or drill chuck
and be supported so that the holder is always square and directly over the hole.
To cut successfully, taps must start square while cutting the first threads.

\bigskip
\textit{A standard "T"\ handle tap wrench and a tap wrench with a pilot shaft
for holding in a collet or drill chuck.}
\bigskip

\secrel{Tapping threads on big CNC machines}

Automatic tapping on a modern CNC milling machine is accomplished by reversing
the spindle. The tap is held in rigid holders and the spindle is electronically
synchronized with the feed to turn in relation to pitch of the tap. It does this
so accurately, a "floating"\ holder is not necessary. It is amazing to see a
part being tapped at 2000 rpm! Metal will cut better at a higher rpm and the
quality of the tapped holes will be better. Of course, you can't do this on
inexpensive equipment, but I thought you might find it interesting.

\secrel{Tapping heads for manual mills and drill presses}

What you may be able to afford some day is a small tapping head. They can be
used with manual drill presses and mills. A good quality tapping head costs
about \$400 (1998). They are designed to turn in one direction when the pressure
is applied to drive the tap into the part and in the opposite direction to
unscrew the tap when the tapping head is raised. The tapping head is usually
geared to turn the tap at twice the input rpm when backing the tap out of the
work. This makes the process more efficient. They make several models, and for
small taps I prefer a cone clutch design over the dog clutch. Dog clutches
engage with a "snap"\ which has a tendency to break small taps. Also, they
cannot tap left-handed threads. Cone clutches operate more smoothly.

When a tapping head is used, the operator must put enough pressure on it to
start the threading process. The tap will then be screwed into the hole at the
rate of its pitch. If the feed rate, controlled by the operator, does not equal
the pitch of the tap, the clutch will disengage. On a drill press, the depth of
the tapped hole will be controlled by the "depth control stop". When you are
tapping blind holes, you set the stop to make sure the tap will not go all the
way to the bottom. Measure the distance to go and reset the stop to eliminate
any error. A tap will go deeper than the stop is set because it has to disengage
the clutch. When the tap is unscrewed, a slight withdrawal pressure is needed to
operate the opposite clutch and any excess pressure will "tear"\ the thread as
the tap leaves the now-threaded hole. A definite feel for the process is needed
to operate these labor saving devices effectively.

\secrel{Tricks of the trade\ldots A tapping tip from Bob Shores}

I have read many tips on tapping holes\ --- some good, some not. Five years ago
I dreamed up a tapping method for small holes. I tap a lot of holes with 0-80
and 2-56 threads, and since I have been using this method, I have not broken a
tap in five years.

After drilling the hole in your part to the proper size, the drill bit is
removed from the chuck without disturbing the work, A 2"\ aluminum disk, knurled
on the outside and drilled and tapped for a 4-40 hex bolt grips the tap just
above the flutes. The end of the tap is gripped in the drill chuck and lowered
until it just touches the work. The chuck is then loosened to allow the tap to
turn freely. The disk holding the tap is turned with your thumb and forefinger.
The drill chuck acts as a guide to keep the tap running true, and your fingers
are very sensitive to the amount of torque being applied. To break a tap you
would have to apply a lot of force.

\bigskip
--- Bob Shores
\bigskip

HOLE SLIGHTLY LARGER THAN TAP

2"\ DIA $\times$ 1/4"\ KNURLED ALUMINUM DISK

4-40 $\times$ 1"\ HEX BOLT

DRILL CHUCK

7"\ KNURLED ALUMINUM DISK

TAP
\bigskip

\secup

\secrel{6.3 LATHE CUTTING TOOLS}\secdown

\secrel{Right- and left-hand tool shapes}

Again, the difference between a lathe and a mill is that the work turns on a
lathe and the tool turns on a mill. Before we get into grinding lathe tools
let's define this type of cutting tool as a single point tool that is fixed
(doesn't cut by rotating) and held in a tool block on the crosslide of a lathe.
By the way, lathe tools can be cut as "right-hand"\ or "left-hand"\ shapes. The
reason we call a tool a "right-handed"\ tool when the cutting edge is on the left
is because it is designated by which way the chip leaves the cutting tool. A
left-handed tool is designated as such because the chip will go to the left as
it cuts. The cutting edge will be on the right. The standard tool is a
right-hand tool, and a right-hand ground tool bit is included with each Sherline
lathe to help get a novice started. Scissors or sheet metal shears are also
defined this way, because the cutoff falls to the left or right. Right-handed
people will usually prefer "left-hand"\ shears.

\bigskip
CHUCK

TAILSTOCK

LEFT-HAND TOOL

RIGHT-HAND TOOL

\bigskip

\textit{A tool is called a "left-hand tool"\ because the chip comes off to the
left even though the cutting face is on the right. On a right-hand tool the chip
comes off to the right.}

\secrel{Why you should ovoid cheap carbon steel tools}

The Sherline lathe has been designed to use 1/4"\ square tool bits. You don't
have many choices when it comes to grinding your own lathe tools. The shapes you
will need to produce special parts can't be made with standard cutting tools. If
you read an old book on machining, it may have mentioned carbon steel tools, and
these are case hardened tools. The only way you might find these inferior tools
would be in cheap imported sets. Today, the labor and machine cost to produce a
good cutting tool exceeds the material cost by so much that it just doesn't make
sense to use cheap steels.

\bigskip
\textit{A selection of 1/4"\ high speed steel and carbide tipped cutting tools
for miniature machining. Shown are left to right: A high speed steel boring
tool, right-and left-hand cutting tools, right- and left-hand carbide tools and
a carbide $60^o$ threading tool.}

\secrel{High speed steel tools}

For the average machinist, "high speed steel"\ will always have a use in a
machine shop. It is inexpensive and easy to grind and shape. High speed steels
comes in a variety of grades. For the average work done in a home shop, M-2 tool
steel is more than adequate. M-5 would be considered top-of-the-line. Some tool
steels contain several other metals to add to their life. Cobalt is a common
additive that makes the cutting edge less prone to "chipping"\ and is effective
in adding life to a tool. Usually, the more high speed tool steel costs, the
harder it will be to grind and the longer it will hold an edge. You will find
the answers to all engineering problems are a trade-off, just as this one is.

\secrel{Brazed carbide and diamond tipped tools}

Sherline offers brazed carbide tools in left, right, and $60^o$ threading.
Brazed carbide tools have a shorter life than inserted tip carbide tools because
the carbide has been brazed to the holder and their different expansion rates
can causes problems. The left-hand brazed carbide tool works well with the
flycutter and is included with its purchase. They can be useful but you need a
diamond wheel to resharpen them. One thing you must realize is that some
materials can only be cut with carbide. You don't have a choice. These materials
usually have an abrasive nature and are not that "hard". In general, hard
materials (heat treated tool steels) can't be successfully machined. You can't
chuck up an end mill in a lathe and machine its shank down with carbide or
diamond cutting tools. Diamond cutting tools are used to get beautiful finishes
on nonferrous materials and shouldn't be used on steels. The carbon in the
diamond will weld to the carbon in the steel and destroy the diamond.

Heat treated tool steels are shaped by grinding, not by turning. Cutting the
silver and copper material used in the armature of an electric motor would be a
good application for a diamond tipped tool.

\bigskip
INSERT TIP

TOOL

TOOL POST (P/N 7600)

SPECIAL TORX DRIVER FOR TIGHTENING INSERT HOLD DOW SCREW (INCLUDED)
\bigskip

\textit{Carbide insert tool and tool post}
\bigskip

\secrel{Inserted tip carbide tools}

I would first like to emphasize that I believe the prime method for cutting
metal on a miniature lathe should be high speed steel cutting tools. They are
inexpensive, easy to sharpen and can be ground to make "form"\ tools. On the
other hand, insert tools are expensive and can't be resharpened or shaped, but
there are times they can be a lifesaver. In some applications they are the
perfect tool for the job.

The obvious difference between brazed carbide tools and inserted carbide tools
is that the tip is held on with a screw rather than brazed onto a piece of
steel. This fact has a lot to do with the success of insert tools in recent
years. Steel and carbide have slightly different expansion rates which can cause
premature failure of the carbide tip. It is surprising that a small screw will
hold these inserts tight enough to accurately cut metal, but it does. We run 20
horsepower computer controlled lathes at Sherline that can remove metal at a
rate of 2 pounds (1 kg.) a minute with these tools and have few failures.

The reasons I believe insert tools should have a place in your shop is they are
ready to use, they hold their cutting edge when cutting exotic metals or
abrasive materials and they can speed up the cutting process.

\secrel{The rules on cutting speeds are different for inserted tip tools}

Normal cutting speed rules don't apply to the same extent as when using high
speed steel. Stainless steel can be cut at triple the rate with these tools
compared to high speed steel tools. This higher RPM also puts you in a better
horsepower range on a small motor. Another interesting fact is that you can get
a better finish on some steels, such as cold rolled, by turning up the RPM.
Insert tools don't need cutting oils to work well, but I still use a few drops
now and then. The lack of messy cutting oil can be an important factor when
working on your kitchen table, as it keeps the work area cleaner.

\secrel{Positive and negative rake tools}

I experimented with various tools before making a choice. I wanted a cutting
tool that had a positive rake. I don't believe the tools we manufacture are
rigid enough for the normal negative rake tools which cost less and are far more
popular.* These tools are designed for machines that weigh hundreds of times
more than a Sherline Lathe. Positive rake tools have to be sharpened at the time
of manufacture which adds to their cost, but also adds to their performance on a
Sherline lathe. You can't use any insert in these holders unless it has a
positive cutting edge.

\bigskip
*NOTE: Sherline also offers a negative rake tool holder designed especially for
use on the Sherline lathe which allows the use of a $55^{o}$ negative rake tip.
Because of the design of the holder, this tool cuts like a positive rake cutter.
This gives you the best of both worlds. A description of that tool can be found
on the next page.

\bigskip
CUTTING EDGE

NEGATIVE RAKE TIP

CUTTING EDGE

POSITIVE RAKE TIP
\bigskip

\textit{Negative and Positive Rake Carbide Cutting Tips. Negative Rake tips can
be held upside down giving 4 cutting edges. Positive Rake tips cut from one side
only, but cut better.}
\bigskip

Another choice I made was the .015"\ (4mm) radius on the tip. A large tool radius
can give good finishes on a full size machine, but it can cause havoc on small
diameter parts or miniature machines. Large radius tools create high tool loads
because of their large cutting surface. On the other hand, a radius smaller than
.015"\ will chip too easily.

The difference between the $80^{o}$ and $55^{o}$ tools is that the $80^{o}$ tool
is a little stronger at the tip, but the $55^{o}$ tool can get into corners
better. If you plan to only purchase one, buy the $55^{o}$. The $80^{o}$ tool is
a good choice to use for rough cuts.

\secrel{Reading the chips}

To get maximum life out of these tools be sure to increase the "feed"\ as you
increase the RPM. The chip should have a tight curl to it and break off into
short lengths. These chips can be very HOT, Remember, your hands are closer to
the cutting edge when using miniature machine tools, so use caution.

\secrel{Cutting harder surfaces}

Another plus for insert tools is their ability to cut hard or abrasive
materials. Don't plan on machining hardened tool steel with them, but you can
cut through a work-hardened surface on stainless steel with ease.

Carbide tools can simplify many machining operations, but will never solve
problems caused by poor machining practices. As with all machining operations,
WEAR EYE PROTECTION.

\bigskip
\textit{The negative rake tool holder is shown with a box of spare carbide tips
and the tightening wrench. The insert is angled down slightly which provides
clearance and is supported all the way down to the table, which provides support
for stiffness.}

\secrel{The 3/8"\ IC $55^{o}$ negative rake tool holders}

I believe Sherline's 3/8"\ IC $55^{o}$ negative rake indexable holder will bring
a lot of enjoyment to your machining, particularly if you have trouble grinding
good tools or if you choose to turn difficult materials such as stainless steel.
The indexable carbide insert sits on the tool holder at a $5^{o}$ negative
angle. This gives the sides of the cutter clearance even though the insert has
square sides. By having square sides, both the top and bottom of the insert can
be used as cutting edges. This gives you front and back, top and bottom of both
sides for a total of four cutting edges on each insert. Though not inexpensive,
when you consider you are getting four cutting tools in one, it is really a
pretty good deal.

Remember that carbide cutting tools are a little more brittle than high speed
steel and take care not to break the insert. If you break a chip out of one
surface, I don't recommend that you use the cutting surface on the other side.
The insert would not be properly supported on the tool holder with material
missing from the lower surface which should be resting solidly on the tool
holder. Damage to the tool holder can occur if you use a broken insert in this
fashion.

\secrel{Precautions for making deep cuts}

Keep in mind also that it is not good machining practice to feed a tool straight
into a part any further than necessary. Parting-off tools are designed for this
task, but regular cutting tools are not. Going straight in puts two cutting
surfaces to work at once, and as you get deeper and deeper into the part, you
can overload the cutting capacity of the machine and cause it to jam. If you
must feed straight in to cut a groove, for example, go a short distance and then
open up the slot side to side, using only one cutting surface of the tool at a
time. Then go a little deeper and repeat the side to side cuts.

\secrel{Don't be afraid to crank up the RPM a bit}

An advantage of this tool is that it will cut in either direction. It can also
give good finishes on hard-to-machine materials such as cold rolled steel.
(Note: the best finishes on soft materials such as aluminum, brass or leaded
steels will still be achieved using a good, sharp high speed steel tool. However
the carbide insert tool will still do a very good job and will last almost
forever on these materials.) A good finish on harder materials using a carbide
tool is accomplished by turning up the RPM of the spindle about three times
faster than if you were using a high speed steel tool. In fact, when making most
cuts with this tool, don't be afraid to turn up the RPM and feed the tool
rapidly. Of course, you must have the part you are machining held in a setup
sufficiently secure to accomplish this.

\secrel{Design of the negative rake tool holder}

The holder is manufactured from 7075 Aluminum, which is approximately twice as
hard as regular aluminum in a T6 condition. This material also costs
approximately twice as much, but I felt it would be money well spent to insure a
long life for the holder.

\secrel{Design of the insert cutter}

The carbide insert is designed in such a way that it cuts like a positive rake
cutter. Positive rake cutters don't require as much rigidity as negative rake
cutters. This type of design allows the advantages of a negative rake cutter
(four cutting edges per insert) without requiring the rigidity that can't be
found in bench type machines.

Replacement inserts for this holder are available both from Sherline and from a
number of tool manufacturers. They are usually designated DNMG-331.

\bigskip
NOTE: ALWAYS USE CUTTING OIL WHEN USING THE CUTOFF TOOL

\secrel{The cutoff or "parting"\ tool}

After completing a part in the lathe it is frequently necessary to separate the
part from the excess material used for chucking. This operation is best
accomplished with the use of a cutoff tool or "parting tool"\ as it is sometimes
called. The Sherline cutoff tool and holder consists of a very slender high
speed tool steel cutting blade mounted in a special tool holder. The thinness of
the blade (.040") enables it to feed into the part quite easily and at the same
time minimizes the amount of waste material. One word of caution; never use a
parting tool on a part mounted between centers. The part may bind on the cutter
and result in a scrapped part or a broken cutting tool.

\secrel{The proper way to cut off a part}

Always try to lay work out so the cutoff tool is used as close to the spindle as
possible. Set blade height by sliding the blade in its slot in the tool holder.
It should be set so the tip is at the same height as the centerline of the part
being cut. An unusual part diameter may require a shim to be placed under the
front or rear of the holder to accomplish this.

NOTE: ALWAYS USE CUTTING OIL WHEN USING THE CUTOFF TOOL. The cut will be made
much smoother, easier and cooler.

The turning speed for parting should be approximately one half the normal
turning speed for any given material, and the feed rate should be a little heavy
so the chip will not break up in the slot. If speed and feed are correct, there
will not be any chatter, and the chip will come out as if it were being
unrolled. Coolant (cutting oil) plays a major roll in this occurring properly.

\bigskip
HEIGHT OF TOOL TIP SHOULD BE SET TO CUT AT CENTERLINE OF PART
\bigskip

\textit{A parting tool is used to separate a completed part from its raw
material workpiece.}

\bigskip

If the tool chatters, first check to see if the work is being held properly.
Then decrease speed (RPM) or increase feed rate or both. Once the blade has
chattered, it leaves a serrated finish which causes more chatter. Sometimes a
serrated finish can be eliminated by stopping the spindle, adding a liberal
amount of cutting oil, bringing the blade up so there is a slight pressure on it
without the spindle turning. and then turning by hand or as slowly as possible
with the speed control.

Very small work may be completely cut off when held in a chuck and allowed to
fall onto the crosslide. It is too small and light to cause any damage. Hollow
articles such as rings may be caught on a piece of wire whose end is held in a
suitable position.

\secrel{SHARPENING INSTRUCTIONS}

To sharpen the blade, use the tool support on the grinder set in such a way that
it will produce a $7^{o}$ to $10^{o}$ angle on the blade (lop to bottom). (See
Figure 1 below.)

\bigskip
FIGURE 1 --- Side view of blade.

$7^{o}$ to $10^{o}$

FIGURE \ --- Top view of blade (enlarged) when ground for ``parting off''

SIDE VIEW

TOP VIEW
\bigskip

If you are sharpening the blade to "part off, the blade should have an
additional angle of approximately 5o when viewed from the top with the point on
the right. (See Figure 2.) Normally the angle would be as high as 15o but the
.040"\ thickness of the blade would not be rigid enough and the blade could bend.
If you want to cut grooves, don't put any angle on the blade when seen from the
top.

\secrel{Mounting the cutoff tool using o rear mounting block}

The rear mounting block is a simple spacer block that allows you to mount the
cutoff tool and holder to the table on the back side of the part. Because the
part is rotating "up"\ on the back side, the tool must be flipped over in the
holder. The mounting block raises the tool holder the amount needed to put the
tip of the tool back at the right cutting height. This will save you time by
being able to leave the cutoff tool holder mounted to the table while you use
the regular tool post in its normal position on the front side of the part.

\bigskip
USE LONGER 10-32 $\times$ 1-3/4"\ SOCKET HEAD SCREW PROVIDED

P/N 3002 CUTOFF TOOL AND HOLDER

PART

STANDARD TOOLPOST AND CUTTING TOOL

LATHE CROSSLIDE
\bigskip

\textit{FIGURE 3\ --- Seen looking from headstock toward tailstock. (Arrow shows
direction of part rotation.) The rear mounting block is shown in position under
the cutoff tool holder and mounted to the "back"\ side of crosslide table. The
cutoff tool holder can now be left mounted to the table, ready for a parting
operation at any time. There is no need to remove the standard tool post in
order to "part off the work.}
\bigskip

The mounting block is placed between the standard cutoff tool holder (P/N 3002)
and the lathe crosslide. It is mounted on the back side of the part, or me side
away from the crosslide handwheel, The longer 10-32 x 1 -3/4"\ socket head screw
provided with the rear mounting block is used to attach the unit to the
crosslide table. (Use the T-nut that came with the cutoff tool holder.) Note
that the hole in the mounting block is not in the center. Rotate the block to
find the position where the sides line up with the sides of the cutoff tool
holder.

Loosen the two clamping screws which hold the cutoff tool blade in place. Turn
the blade over so the cutting tip is facing down and mount it as shown in Figure
3. Adjust the tip of the tool to the desired height by sliding it back and forth
in its slot. Lock it in position with the two clamping screws.

Even though the tool is upside down, all the same rules about its use still
apply. The only difference is that the crosslide table is now cranked toward the
operator to cut off a part.

\bigskip
\textit{A parting tool is held upside down in the tool holder which is mounted
to the rear mounting block. This simple accessory is helpful if you are doing
repeated cutoff operations and don t want to keep switching toolposts.}
\bigskip

\secup

\secrel{GRINDING YOUR OWN LATHE TOOLS}\secdown

As a home shop machinist you are going to have to shape tools the old fashioned
way, and that means you need a grinder. A bench grinder doesn't have to be
expensive to work well, but it does require good "wheels"\ for high speed steels.
Try to find a source for grinding wheels from an industrial supplier. Some of
the wheels that come with inexpensive grinders wouldn't sharpen a butter knife.
Sixty grit is a good place to start. A wheel dresser is also a necessity. They
cost less than \$10 and are readily available from good hardware stores.

\secrel{Using a wheel dresser to keep your wheel sharp}

Grinding wheels should be considered cutting tools and have to be sharpened. A
wheel dresser sharpens by "breaking off the outer layer of abrasive grit from
the wheel with star shaped rotating cutters which also have to be replaced from
time to time. This leaves the cutting edges of the grit sharp and clean.

\bigskip
\textit{A wheel dressing tool and spare "star"\ cutter}
\bigskip

A sharp wheel will cut quickly with a "hissing"\ sound and with very little heat
by comparison to a dull wheel. A dull wheel produces a "rapping"\ sound created
by a "loaded up"\ area on the cutting surface. In a way, you can compare what
happens to grinding wheels to a piece of sandpaper that is being used to sand a
painted surface; the paper loads up, stops cutting, and has to be replaced.

\secrel{Setting up your grinder and tool rest}

For safety, a bench grinder should be mounted to something heavy enough so it
will not move while being used. The tool support must be used and should be set
at approximately $7^{o}$. Few people have the skill to make tools without a tool
support and in essence it's wasted effort. Tool supports are usually made up of
two pieces that allow you to set your tool rest above or below center. It really
doesn't matter whether it's above or below as long as the support is at $7^{o}$.

The reason tool supports are designed like this is so they can be used for a
variety of uses, not just tool bits. What this means is that if the tool support
is above or below center, it must be adjusted as the wheel diameter changes.

\bigskip
GRINDING WHEEL

TOOL REST
\bigskip

\textit{FIGURE 4 — Set tool rest at any height, but at $7^{o}$ angle from centerline
of wheel.}
\bigskip

\secrel{Getting started on making a toot}

Now it's time to make a tool, and whether you turn the job into a major project
is up to you!

\begin{framed}
CAUTION!

When working around grinders it is an absolute necessity to wear EYE PROTECTION.
Grinding debris is thrown out at high velocities and can damage not only eyes,
but also expensive glasses. Wear safety glasses or a full face shield.
\end{framed}

If you've never sharpened a tool, take a close look at how ours are sharpened.
Let's duplicate the right hand tool on the opposite end of the blank. Be careful
you don't cut yourself on the blank or the sharpened end while working with it.

First dress the wheel by taking the dresser and setting it on the tool support
square with the wheel and while applying a light pressure move the dresser back
and forth with the grinder running. Unless the wheel is in bad shape, it should
be ready to use in a few passes.

\secrel{Grinding side 1 of the tool}

Turn off the grinder and set the tool support for approximately $7^{o}$ if you
haven't done it yet. If you're not good at guessing at angles use a presharpened
SHERLINE tool to set the angle. Metal cutting tools are very tolerant on angles.
I've always found wood cutting tools more difficult to sharpen. Too little angle
and the "heel"\ of the tool will rub, loo much angle will cause the tool to "dig
in"\ and chatter.

\bigskip
TIP

HEEL
\bigskip

\textit{FIGURE 5\ --- Heel of tool}

Have a cup of water handy to cool the tool with, set the blank on the tool rest
and start grinding side I.

\bigskip
WINDING WHEEL

TOOL REST
\bigskip

\textit{FIGURE 6\ --- Grinding Side 1 of the tool}

\bigskip
NOTE: Became of T angle on laol rest, side of tool is actually cur first.

TOP OF TOOL
\bigskip

\textit{FIGURE 7\ --- Properly ground side I}
\bigskip

Move the blank, back and forth across the face of the wheel until you have
ground a 10$^{o}$ angle on approximately 3/16"\ (4mm) of side 1.

This is where the "positive approach"\ comes in. Unless you push the tool into
the wheel with enough pressure, the tool will bounce around and you'll never get
a good flat cutting surface. It isn't necessary to worry about getting the tool
too hot. Modem day tool steels don't anneal and a little discoloration doesn't
affect the tool life in tool room use. What you should worry about is not
burning yourself or grinding the tips of your fingers off! Concentrate on
holding the 10$^{o}$ angle while moving back and forth. We'll give this edge a
final sharpening later, but now it's time for side 2.

\bigskip
TOP OFTOOL

LESS THEN $90^o$
\bigskip

\textit{FIGURE 8\ --- Grinding side 2 of the tool}
\bigskip

\bigskip
PART

TOOL
\bigskip

\textit{FIGURE 9\ --- Properly ground tool cutting into a corner.}

\secrel{Grinding tide 2 of the lool}

The reason angle B is ground less than $90^o$ is to allow the tool to get into
corners.

Side 2 is ground the same way as side 1. moving the tool back and forth until
you have a point. After you get side 2 ground, cool the tool in the cup of
water.

Now 1 want you to learn another aspect of tool grinding. It is important to know
when you have ground the surface up to the cutting edge, especially when
resharpening lathe tools. Take the tool you just ground and bring it up to the
wheel at a slightly different angle than you just ground for this experiment.
Watch the point that touches the wheel first and you will notice that the sparks
will bounce off the cutting edge only where the wheel has ground from top to
bottom.

\bigskip
SPARKS

TIP

SPARKS AT TIP OF TOOL
\bigskip

\textit{FIGURES 10A\ --- Tip not yet ground flat and 10B, Tool ground flat all
the way to the tip.}
\bigskip

This tells you when the tool has been sharpened without taking it away to look
which allows you to grind flat and true surfaces. If you sharpen a tool for a
Sherline lathe, use a 1/4"\ square tool blank and keep the cutting edge up to
the top of the blank; the tool will come out on center without shims. You will
have to be precise grinding the third side to accomplish this.

\secrel{Grinding side 3}

\bigskip
ROTATE TOOL

APPROX $15^o$

GRIND UNTIL SPARKS JUST REACH TIP OF TOOL
\bigskip

\textit{FIGURE 11\ --- Grinding the "Hook"\ into side 3}
\bigskip

Now you will use the skill you have developed grinding the second side. Set the
blank on the support with the $10^o$ (side 1) up. The tool has to be brought up
to the grinding wheel with a slight angle so you don't grind the tip below
center. With the tool setting on the rest, move the tool in and grind until you
see sparks bouncing off the cutting edge where the corner of the wheel is lined
up with the back part of the $10^o$ face. When this happens, slowly decrease the
angle without pushing the tool in any more until sparks bounce all the way to
the tip. Stop as soon as this happens. You may inspect it, and the surface
should be entirely ground. The recommended way is to put more "hook"\ on the tool
than I have suggested, but I have found that the slight increase in performance
is offset by the problems encountered resharpening these tools.

\bigskip
SIDE

FRONT
\bigskip

\textit{FIGURES 12A\ --- Normally recommended "hook"\ ground into tool and 12B,
Simpler method suggested for Sherline tools.}
\bigskip

To put the finishing touches on your tool, you have to "kiss off"\ sides 1 and 2
again. You must carefully line up side 1 with the wheel and bring it to the
wheel in a positive manner with very little pressure. Then watch for the sparks
on the cutting edge. What you're trying to accomplish is to make the tool set
against the wheel on the same plane as when you first ground side 1. If the tool
is held too rigid, it will not align itself, too loose and it will bounce
around.

\secrel{"Breaking"\ the point}

Use the same method on side 2. The tool should be ready to use except for the
point. I always put about a .010"\ (.2mm) "break"\ on the point by holding the
tool with the point aimed at the wheel face. Because two angles converge at the
point, the angle in relation to the sides is greater. Think about it!

\bigskip
ANGLES APPROX EQUAL

TOP OF TOOL
\bigskip

\textit{FIGURE 13\ --- Putting a .010"\ "Break"\ on the tip of the fool.}
\bigskip

This means that if you set the tool flat on the tool rest, the tool rest angle
would have to be increased to get an even flat. This wouldn't be worth the
effort. so the easy way is to freehand it. 1 always start by touching the heel
of the tool first, and then change the angle until a slight flat is put on the
tip. Of course, the angle you're holding it at has to be close when starling to
get desired results.

\bigskip
\textit{FIGURE 14\ --- Hand holding the tool to "Break"\ the point saves
resetting the angle on the tool rest.}
\bigskip

The purpose of this flat is to improve finish and tool life. I don't recommend a
large radius on the tip of tools used on small machines. These machines are not
rigid enough to get the desired results from this practice and cause "chatter"
problems.

\secrel{Some final tips}

The finished product should be a right handed tool, have flat cutting surfaces
(except for the radius caused by the wheel), have a slight flat on the tip, and
a tip angle of less than 90$^o$. Tools used on lathes such as the Sherline will do
all their cutting at the tip of the tool because they don't have the horsepower
for 1/4"\ (6mm) cuts.

I don't recommend using oil stones to improve the edges. After a few minutes use
with an occasional dab of cutting oil a properly sharpened tool will hone itself
in.

I always believe the final sharpening to a tool should take place with the wheel
cutting the cutting edge of the tool from the top of the tool to the bottom when
using bench grinders.

I realize I've given a great deal of information on how to do what I call a
"simple operation", but these are very difficult instructions to write because
I'm trying to tell you how to control your hands, not a simple machine.

\bigskip
SIDE VIEW

BOTTOM VIEW
\bigskip

\textit{A typical boring tool}
\bigskip

\secrel{Grinding boring tools}

Boring tools are the most difficult tools to grind. They should always be made
as rigid as possible. Tool angles around the "tip"\ can be the same as any
cutting tool, but clearances of the tool body have to be considered carefully. A
tool ground with enough clearance for a finished hole may not have enough
clearance to start with when the hole has a smaller diameter. If you have to
bore a hole in a part that has a lot of work in it, have a tool ready to use
that's been checked out on a piece of scrap.

\secrel{Grinding "form"\ tools}

Form tools are used to create a shape that is a mirror image of the shape of the
tool. To grind form tools, a pattern of the finished shape should be at hand and
you should have a workable design for the tool in mind. You must be able to
achieve the shape you want with the tools you have to work with. For example,
you can't grind a 1/8"\ (3mm) groove 1/4"\ (6mm) deep into your tool with a 1/2"
(12mm) wide grinding wheel.

Intricate form tools are usually made by tool and cutter specialists who have
expensive shop rates and use precision grinders, diamond dressers and the large
variety of wheels available to them. However, all is not lost if we have a good
pair of hands with a good mind driving them! We can use the grinding wheel
corners on our \$50 grinder and generate the shape one half at a time on each
side of the tool and still get our job done.

\bigskip
SHAPE OF PART DESIRED

STEP 1
                     
STEP 2

FORM TOOL
\bigskip

\textit{Drawing A shows a Typical Form Tool made by a custom toolmaking shop.
Drawing B shows a home shop method of achieving the same finished shape in two
steps with a tool that can be ground on a bench grinder.}
\bigskip

Form tools don't need any top relief (hook) to work. Use low spindle RPM and
steady feed rates to prevent chatter. The width of a form tool should never
exceed three times the smallest diameter of the finished part.

Like any skill, tool grinding is one that has to develop with time. It is also
the skill that allows you to go one step beyond the average "hacker". Working
with sharp tools makes machining much more pleasant too, so this is a skill
worth spending some time developing. In the long run you will also save a lot of
money by not having to continually buy new cutting tools or pay someone to
sharpen them for you.

\bigskip
\textit{Sparks fly as the finishing touches are put on a lathe tool using a
grinding wheel. Learning to grind good cutting tools is a key to getting good
finishes on your parts.}
\bigskip



\secup

\secup
