\secdown
\secrel{Dedication}\secdown

\textit{Carl Hammons\ --- 1936--1997}
\bigskip

Carl Hammons, my friend and business partner for thirty years, died September
11, 1997 as I was writing this book. We shared thousands of lunches and coffee
breaks over the years we worked together, and much of the knowledge I have
passed on in this book came from Carl. Carl and 1 shared the rare distinction of
having been partners not just once, but twice. We both played different roles in
putting together the product line, and without him it just isn't going to be as
much fun.

When we joined forces for the second time, we had an agreement that eliminated
any need to financially justify the purchase of a new piece of equipment. We
would buy machines that interested us and find a job for them later. The laser
engraver was a perfect example of this, but now we couldn't get along without
it. It may seem contrary to smart business practice, but that's the way we did
it. I have no regrets, for we were always the happiest when we were confronted
with a new set of technical problems. Therefore, I dedicate this book to Carl
Hammons; my business partner, my friend.

\bigskip
I should also credit the English teachers in the Cranston, Rhode Island school
system for forcing a not-so-willing student enrolled in the "boys general class"
to learn enough about our language to dare to take on the task of expressing
difficult concepts in simple words. I graduated in 1953. You, the reader, will
be the ultimate judge of their (and my) success in this undertaking.

\bigskip
--- Joe Martin

\bigskip
\textit{The photo composition above is a joint effort. The photo of Cart was
taken by his wife Barbara. The photo of Swan Lake, Montana, a favorite spot of
Carl's, was taken by friend Wayne Armstrong. The two images were composed in
PhotoShop by artist Elaine Collins}

\secup
\secup
